### 2017-04-05 Chapters/theory.tex
Some small frenchproofreading and correct overfull/underfull hbox when possible

\chapter{Dynamical Allocation of Cellular Resources as an Optimal Control Problem: Novel Insights into Microbial Growth Strategies}
\chaptermark{Dynamical allocation of cell resources}
\label{chap:theory}

@@

Ce chapitre est dédié à une approche théorique du problème d'allocation des ressources cellulaires chez les microorganismes.
En Section \ref{sec:model}, nous modélisons la cellule comme un auto-réplicateur devant allouer une ressource commune (les nutriments du milieu) à deux secteurs distincts de macromolécules: la machinerie métabolique qui extrait les nutriments de l'environnement et les convertit en précurseurs utilisables, et la machinerie d'expression génique qui utilise ces précurseurs pour produire de nouvelles macromolécules (Fig.~\ref{fig:self_replicator}).
Ce modèle très simple comprenant 2 réactions (Eq.~\ref{eq:reactions}) se traduit mathématiquement par deux équations différentielles ordinaires (Eqs~\ref{eq:pdef} et \ref{eq:rdef}) faisant intervenir le paramètre $\alpha$ qui va devenir clé dans notre étude.
$\alpha$ représente la proportion massique des précurseurs qui sont utilisés pour produire de la machinerie d'expression génique, au détriment de la machinerie métabolique (qui elle, est produite à la proportion $1-\alpha$).
Comme on l'a déjà vu avec les lois de croissance établies à l'état stationnaire (voir Chapitre~\ref{chap:introduction}), cette proportion varie avec l'environnement, on s'attend donc à ce que $\alpha$ en fasse autant, augmentant d'autant que l'environnement devient riche (Fig.~\ref{fig:scott_rnaprot}).
Cependant, de quelle manière varie-t-il entre deux états de croissance stationnaires, lorsque les composants cellulaires sont en déséquilibre et que la cellule doit adapter sa composition ?

Comme nous l'avons vu dans le Chapitre~\ref{chap:introduction}, les lois de croissance à l'état stationnaire s'expliquent facilement si l'on considère que la cellule maximise son taux de croissance.
En Section~\ref{sec:growthlaws}, nous voyons en effet que pour chaque environnement, un $\alpha$ unique existe qui maximise la croissance de la cellule, mais surtout que ce $\alpha^*_{opt}$ est d'autant plus important que l'environnement est riche (Fig.~\ref{fig:model_analysis}).
De fait, les lois de croissance à l'état stationnaire sont parfaitement prédites par notre modèle simple en considérant uniquement que la cellule maximise son taux de croissance (Fig.~\ref{fig:model_validation}).
Peut-on appliquer le même principe, dans un contexte cette fois dynamique, de façon à prédire comment la cellule distribue ses ressources lors d'une transition de croissance ?

Dans ce contexte, $\alpha$ est désormais une fonction du temps.
Nous cherchons donc quelle fonction du temps maximise un objectif donné.
Ce problème, formulé en Section~\ref{sec:optimalcontrolproblem}, est un problème de contrôle optimal, puisque l'objectif à optimiser est une fonction d'une autre fonction, plus communément appelée fonctionnelle.
Nous choisissons comme équivalent dynamique de la maximisation du taux de croissance à l'état stationnaire, la maximisation de la production de biomasse sur un intervalle de temps comprenant un changement abrupt de l'environnement (Eq.~\ref{eq:biomassdyn}).
Nous résolvons ce problème en Section~\ref{sec:optimalcontrolsolution} en couplant l'utilisation du principe du maximum de Pontryagin avec de l'optimisation numérique.
La solution optimale obtenue montre que le contrôle $\alpha$ doit prendre alternativement les valeurs 0 ou 1 jusqu'à ce que le système soit stabilisé sur un nouvel équilibre (Fig.~\ref{fig:optimalcontrol} et Eq.~\ref{eq:optcontrol}).
En d'autres termes, la meilleure production de biomasse s'obtient si, à tout moment, la cellule aiguille 100\% de ses ressources vers un seul secteur, en alternant entre l'un et l'autre jusqu'à ce que le nouvel état stationnaire de croissance soit atteint.
Ce type de solution se rencontre souvent en contrôle optimal, et répond au nom de concept TOR (Tout-Ou-Rien, ou \textit{bang-bang-control} en anglais).
Mais ce contrôle est-il pertinent d'un point de vue biologique ?
La cellule a-t-elle suffisamment d'information à sa disposition pour réaliser une telle transition, laquelle repose sur des changements abrupts de l'expression génique à des instants bien précis ?

@@

Finalement, nous montrons en Section~\ref{sec:strategies} qu'une stratégie plus complexe mesurant à la fois les précurseurs et la concentration en machinerie d'expression génique est tout à fait capable de réaliser un contrôle proche du tout-ou-rien optimal (Eq.~\ref{eq:stratswitch} et Fig.~\ref{fig:onoffresults}).
Mais surtout, en réutilisant un modèle du système ppGpp, connu pour réguler la synthèse des ribosomes chez la bactérie \textit{Escherichia coli}, nous montrons que ce système répond, au moins structurellement, aux exigences d'une telle stratégie de régulation (Fig.~\ref{fig:ppGppsurface}).
Cela procure un nouveau point de vue sur ce système, d'ailleurs largement répandu chez de nombreux microorganismes.
En intégrant l'information provenant à la fois de la quantité de ribosomes et de celle des précurseurs, il permet d'allouer efficacement les ressources de la cellule lorsque celle-ci doit s'adapter rapidement à un nouvel environnement.
\end{chapter_summary}
\selectlanguage{english}

@@

For this reason, growth-rate maximization is a central hypothesis in a number of recent theoretical studies of resource allocation using coarse-grained models of the cell~\cite{molenaar_shifts_2009,scott_interdependence_2010,scott_emergence_2014}.
The models deliberately reduce the molecular complexity of regulatory networks so as to focus on generic explanatory principles~\cite{servedio_not_2014}.
Along these lines, Molenaar \textit{et al.} developed a series of simple models of the microbial cell, taking into account that growth requires the synthesis of proteins playing a role in metabolism (transporters, enzymes) and gene expression (ribosomes), in varying proportions.
Allocation parameters that maximize the growth rate were shown to account, at least in a qualitative way, for the variation of the amount of ribosomal protein as a fraction of total protein in different growth media, and for the occurrence of overflow metabolism above certain growth rates~\cite{molenaar_shifts_2009}.
Using another coarse-grained model of the cell, centered on amino acid supply (metabolism) and demand (protein synthesis), Scott \textit{et al.} derived empirical growth laws with linear relations between the ribosomal protein fraction and the growth rate, in conditions where the nutrient supply or demand are altered~\cite{scott_interdependence_2010,scott_emergence_2014}.
In their model, maximization of growth rate requires maximization of amino acid flux and is achieved for a specific, unique value of the ribosomal protein fraction.
Based on a structurally similar model, Maitra and Dill related optimal resource allocation to the basic constants of the metabolic and gene expression machinery, in particular energy efficiency~\cite{maitra_bacterial_2014}.

@@

Bearing in mind that the major constituents of the cell are macromolecules (DNA, RNA, proteins), produced from precursor metabolites, a fundamental resource allocation question is the following: How much of the cellular resources are invested in the making of new macromolecules (gene expression machinery) and how much in performing other functions, in particular producing metabolic enzymes involved in the uptake of nutrients and their conversion to precursor metabolites (metabolic machinery)?
In order to address this question, we consider a self-replicating system composed of the gene expression machinery ($R$) and the metabolic machinery ($M$).
The system, shown schematically in Fig.~\ref{fig:self_replicator}, is thus defined by two macroreactions which are conveniently written as:
\begin{equation}\label{eq:reactions}
\begin{aligned}
S &\overset{V_M}{\longrightarrow} P, \\
P &\overset{V_R}{\longrightarrow} \alpha R + (1-\alpha) M.
\end{aligned}
\end{equation}
The first reaction, catalyzed by $M$, converts external substrates ($S$) into precursor metabolites ($P$).
The second reaction, catalyzed by $R$, converts precursors into macromolecules ($R$ and $M$).
The resource allocation parameter $\alpha \in [0,1]$ defines the proportion of precursor mass used for making gene expression machinery as compared to metabolic machinery.
We will interchangeably use the symbols $M$, $R$, $S$, and $P$ for the components of the replicators themselves and their total mass [g].
We will denote the rates at which the macroreactions occur by $V_R$ and $V_M$ [g~h\textsuperscript{-1}].

@@

\subsection{A near-optimal feedback control strategy: exploiting information on the imbalance between precursors and the gene expression machinery}
\label{sec:strategies}

@@

As described in the \textit{Results} section, we are interested in control strategies satisfying the following conditions:
\begin{description}
\item[(C1)] The control laws are static functions of the system variables (as opposed to, for instance, functions that depend on derivatives or integrals of the variables).
\item[(C2)] For any given constant environment $E_M$, they drive the self-replicator system towards a unique stable steady state that is not trivial, \textit{i.e.}, with nonzero growth rate.
\item[(C3)] This steady state corresponds to the optimal steady state $(\hat{p}_{opt}^*, \hat{r}_{opt}^*)$, allowing growth at the maximal rate $\mu^*_{opt}$.
\end{description}
It can be directly verified from the functions $f$, $g$, and $h$ defining the nutrient-only, precursor-only, and on-off control strategies (Eqs~\ref{eq:f}, \ref{eq:precursorstrategy}, and \ref{eq:stratswitch}) that they are indeed static functions of the system variables (or the system input, in the case of the nutrient-only strategy). 
Here we show that the other two conditions are also satisfied for all three strategies.

@@

The full specification of the precursor-only strategy demands an expression for the function $g$ in Eq.~\ref{eq:precursorstrategy}.
Recall that Eq.~\ref{eq:alpha_mu_optimal_p} defines $\alpha_{opt}^*$ in terms of the precursor concentration $\hat{p}_{opt}^*$, which leads us to propose the following function $g$:
\begin{equation}
\label{eq:meth_g}
g(\hat{p}) = \frac{\hat{p}}{\hat{p} + \frac{K}{K + \hat{p}}(1+\hat{p})}.
\end{equation}
As shown in \ref{S1_Text} by computing the Jacobian, the system given by Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} and \ref{eq:meth_g} has a single nontrivial stable steady state for any environment $E_M$ (Condition~C2).
Moreover, Eq.~\ref{eq:meth_g} guarantees this steady state to be optimal (Condition~C3).
This can be seen by noting that at steady state, $d\hat{r}/dt=0$ implies $\hat{r}^*=g(\hat{p}^*)$ (Eq.~\ref{eq:rdefnondim}).
In order for the self-replicator to attain a maximal growth rate at steady rate, Eq.~\ref{eq:alpha_mu_optimal_p} needs to be satified, which is the case for the above choice of the function $g$.
Like for $f$, Eq.~\ref{eq:meth_g} is the only choice for $g$ satisfying C1-C3.
\ref{S1_Fig} shows a plot of $g(\hat{p})$ together with a biologically plausible Hill approximation (Eq.~\ref{eq:HillApprox}).

The on-off control strategy is defined in Eq.~\ref{eq:stratswitch} and repeated below:
\begin{equation}
h(\hat{p}, \hat{r}) = 
\begin{cases}
0, \ \textrm{if} \ \hat{r} > g(\hat{p}),\\
1, \ \textrm{if} \ \hat{r} < g(\hat{p}), \\
\alpha_{opt}^*, \ \textrm{if} \ (\hat{p},\hat{r})=(\hat{p}_{opt}^*,\hat{r}_{opt}^*).
\end{cases}
\end{equation}
This strategy drives the system to a single steady state, because the $\hat{p}$-nullcline crosses the function $g(\hat{p})$ only once, as shown graphically in Fig.~\ref{fig:onoffresults}\textit{A}.
In \ref{S1_Text} we argue that this steady state is stable, by taking into account so-called sliding modes on the switching curve~\cite{filippov_differential_1988} (Condition~C2).
Moreover, the steady state coincides with the optimal steady state $(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$ by construction, so that Condition~C3 is satisfied as well.
Fig~\ref{fig:ppGppsurface}\textit{A} shows a plot of $h(\hat{p}, \hat{r})$.

@@

In this section, we derive the main results for the functions $f$, $g$, and $h$ defining the nutrient-only, precursor-only, and on-off control strategies.
For each of these, we prove that the Conditions C1, C2 and C3 from the \textit{Methods} section are satisfied, which we repeat here for clarity:
\begin{description}
\item[(C1)] The control laws are static functions of the system variables (as opposed to, for instance, functions that depend on derivatives or integrals of the variables).
\item[(C2)] For any given constant environment $E_M$, they drive the self-replicator system towards a unique stable steady state that is not trivial, \textit{i.e.}, with nonzero growth rate.
\item[(C3)] This steady state corresponds to the optimal steady state $(\hat{p}_{opt}^*, \hat{r}_{opt}^*)$, allowing growth at the maximal rate $\mu^*_{opt}$.
\end{description}

@@

Ref.~\cite{uter_longrange_2004} provides a typical value for such a reaction, catalyzed by glutaminyl-tRNA synthetase: $k_{\textit{cat,GlnRS}} = 3.2$~s\textsuperscript{-1}, indicating that on average 3.2 glutaminyl-tRNA molecules are produced per glutaminyl-tRNA synthetase molecule per second. 
After conversion to mass units using molar weight from~\cite{freist_glutaminyltrna_1997}, this yields
\[
k_{\textit{cat,GlnRS}} = \frac{3.2 \cdot 147}{64.4 \cdot 10^ 3} \approx 10^{-3} \, \text{g of glutaminyl-tRNA} \cdot \text{g of enzyme}^{-1} \cdot \text{s}^{-1}.
\]
We therefore take
\[
k_M \approx 3.6 \text{ h}^{-1},
\]
and thus obtain an upper bound for $e_M$ in our simulations.

@@

An extremal trajectory is a quadruplet $(\hat{x}(\cdot),\lambda(\cdot),\lambda_0,\alpha(\cdot))$ satisfying Eqs~\ref{eq:model}-\ref{eq:PMP}. The extremal is said to be normal (resp. abnormal) if $\lambda_0<0$ (resp. $\lambda_0=0$). In the normal case, we normalize the adjoint vector so that $\lambda_0=-1$.

@@

\subsection{S1 Table -- Parameter values of self-replicator model}
\manuallabel{S1_Table}{S1~Table}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{./Fig/plot_strategies}
\caption{
\textbf{Simple control strategies for the self-replicator of bacterial growth.}
\textit{A:} Nutrient-only strategy: $\alpha = f(E_M)$.
The dashed, black curve is the (unique) strategy driving the system exactly to the optimal steady state, that is, the state in which growth occurs at the maximal rate supported by $E_M$.
The function $f$ is defined by Eq.~\ref{eq:meth_f} in the \textit{Methods} section of the main text.
The solid, red curve is an approximation of this function by the simple Michaelis-Menten curve of Eq.~\ref{eq:MMApprox}, with $K_{mE} = 1.0$.
\textit{B:} Precursor-only strategy: $\alpha = g(\hat{p})$.
The dashed, black curve is the (unique) strategy driving the system exactly to the optimal steady state.
The function $g$ is defined by Eq.~\ref{eq:meth_g} in the \textit{Methods} section of the main text.
The solid, red curve is an approximation of this function by the simple Hill curve of Eq.~\ref{eq:HillApprox}, with $K_{mp} = 0.06$ and a cooperativity coefficient 2.}
\end{figure}

>>>>

\chapter[Dynamical Allocation of Cellular Resources as an Optimal Control Problem]{Dynamical Allocation of Cellular Resources as an Optimal Control Problem: Novel Insights into Microbial Growth Strategies}
\chaptermark{Dynamical allocation of cell resources}
\label{chap:theory}

@@

Ce chapitre est dédié à une approche théorique du problème d'allocation des ressources cellulaires chez les microorganismes.
En Section \ref{sec:model}, nous modélisons la cellule comme un auto-réplicateur devant allouer une ressource (les nutriments du milieu) à deux secteurs distincts de macromolécules: la machinerie métabolique qui extrait les nutriments de l'environnement et les convertit en précurseurs utilisables, et la machinerie d'expression génique qui utilise ces précurseurs pour produire de nouvelles macromolécules (Fig.~\ref{fig:self_replicator}).
Ce modèle très simple comprenant 2 réactions (Eq.~\ref{eq:reactions}) se traduit mathématiquement par deux équations différentielles ordinaires (Eqs~\ref{eq:pdef} et \ref{eq:rdef}) faisant intervenir le paramètre $\alpha$ qui va devenir clé dans notre étude.
$\alpha$ représente la proportion massique des précurseurs qui sont utilisés pour produire de la machinerie d'expression génique, au détriment de la machinerie métabolique (qui elle, est produite à la proportion $1-\alpha$).
Comme on l'a déjà vu avec les lois de croissance établies à l'état stationnaire (voir Chapitre~\ref{chap:introduction}), cette proportion varie avec l'environnement, on s'attend donc à ce que $\alpha$ en fasse autant, augmentant d'autant que l'environnement devient riche (Fig.~\ref{fig:scott_rnaprot}).
Cependant, de quelle manière varie-t-il entre deux états de croissance stationnaires, lorsque les composants cellulaires sont en déséquilibre et que la cellule doit adapter sa composition ?

Comme nous l'avons vu dans le Chapitre~\ref{chap:introduction}, les lois de croissance à l'état stationnaire s'expliquent facilement si l'on considère que la cellule maximise son taux de croissance.
C'est aussi le cas dans notre modèle, et en Section~\ref{sec:growthlaws} nous voyons que pour chaque environnement, la croissance de la cellule est maximale pour une valeur unique de $\alpha$.
Mais surtout, cette valeur que l'on nomme $\alpha^*_{opt}$ est d'autant plus grande que l'environnement est riche (Fig.~\ref{fig:model_analysis}).
C'est finalement toutes les lois de croissance à l'état stationnaire qui sont prédites par notre modèle simple, sous la simple hypothèse qu'à tout moment, la cellule maximise son taux de croissance (Fig.~\ref{fig:model_validation}).
Peut-on appliquer le même principe, dans un contexte cette fois dynamique, de façon à prédire comment la cellule distribue ses ressources lors d'une transition de croissance ?

Dans ce contexte, $\alpha$ est désormais une fonction du temps.
Nous cherchons donc quelle fonction du temps maximise un objectif donné.
Ce problème, formulé en Section~\ref{sec:optimalcontrolproblem}, est un problème de contrôle optimal, puisque l'objectif à optimiser est une fonction d'une autre fonction, plus communément appelée fonctionnelle.
Nous choisissons comme équivalent dynamique de la maximisation du taux de croissance, la maximisation de la production de biomasse sur un intervalle de temps comprenant un changement abrupt de l'environnement (Eq.~\ref{eq:biomassdyn}).
Nous résolvons ce problème en Section~\ref{sec:optimalcontrolsolution} en couplant l'utilisation du principe du maximum de Pontryagin avec de l'optimisation numérique.
La solution optimale obtenue montre que le contrôle $\alpha$ doit prendre alternativement les valeurs 0 ou 1 jusqu'à ce que le système soit stabilisé sur un nouvel équilibre (Fig.~\ref{fig:optimalcontrol} et Eq.~\ref{eq:optcontrol}).
En d'autres termes, la meilleure production de biomasse s'obtient si, à tout moment, la cellule aiguille 100\% de ses ressources vers un seul secteur, en alternant entre l'un et l'autre jusqu'à ce que le nouvel état stationnaire de croissance soit atteint.
Ce type de solution se rencontre souvent en contrôle optimal, et répond au nom de concept TOR (Tout-Ou-Rien, ou \textit{bang-bang} en anglais).
Mais ce contrôle est-il pertinent d'un point de vue biologique ?
La cellule a-t-elle suffisamment d'information à sa disposition pour réaliser une telle transition, laquelle repose sur des changements abrupts de l'expression génique à des instants bien précis ?

@@

Finalement, nous montrons en Section~\ref{sec:strategies} qu'une stratégie plus complexe mesurant à la fois les précurseurs et la concentration en machinerie d'expression génique est tout à fait capable de réaliser un contrôle proche du tout-ou-rien optimal (Eq.~\ref{eq:stratswitch} et Fig.~\ref{fig:onoffresults}).
Mais surtout, en réutilisant un modèle du système ppGpp, connu pour réguler la synthèse des ribosomes chez la bactérie \textit{Escherichia coli}, nous montrons que ce système répond, au moins structurellement, aux exigences d'une telle stratégie de régulation (Fig.~\ref{fig:ppGppsurface}).
Cela éclaire d'un jour nouveau le fonctionnement et l'origine de ce système de régulation largement répandu chez de nombreux microorganismes.
En intégrant l'information provenant à la fois de la quantité de ribosomes et de celle des précurseurs, ce dernier est capable d'allouer efficacement les ressources de la cellule lorsque celle-ci doit s'adapter rapidement à un nouvel environnement.
\end{chapter_summary}
\selectlanguage{english}

@@

For this reason, growth-rate maximization is a central hypothesis in a number of recent theoretical studies of resource allocation using coarse-\linebreak grained mod\-els of the cell~\cite{molenaar_shifts_2009,scott_interdependence_2010,scott_emergence_2014}.
The models deliberately reduce the molecular complexity of regulatory networks so as to focus on generic explanatory principles~\cite{servedio_not_2014}.
Along these lines, Molenaar \textit{et al.} developed a series of simple models of the microbial cell, taking into account that growth requires the synthesis of proteins playing a role in metabolism (transporters, enzymes) and gene expression (ribosomes), in varying proportions.
Allocation parameters that maximize the growth rate were shown to account, at least in a qualitative way, for the variation of the amount of ribosomal protein as a fraction of total protein in different growth media, and for the occurrence of overflow metabolism above certain growth rates~\cite{molenaar_shifts_2009}.
Using another coarse-grained model of the cell, centered on amino acid supply (metabolism) and demand (protein synthesis), Scott \textit{et al.} derived empirical growth laws with linear relations between the ribosomal protein fraction and the growth rate, in conditions where the nutrient supply or demand are altered~\cite{scott_interdependence_2010,scott_emergence_2014}.
In their model, maximization of growth rate requires maximization of amino acid flux and is achieved for a specific, unique value of the ribosomal protein fraction.
Based on a structurally similar model, Maitra and Dill related optimal resource allocation to the basic constants of the metabolic and gene expression machinery, in particular energy efficiency~\cite{maitra_bacterial_2014}.

@@

Bearing in mind that the major constituents of the cell are mac\-ro\-mol\-e\-cules (DNA, RNA, proteins), produced from precursor metabolites, a fundamental resource allocation question is the following: How much of the cellular resources are invested in the making of new macromolecules (gene expression machinery) and how much in performing other functions, in particular producing metabolic enzymes involved in the uptake of nutrients and their conversion to precursor metabolites (metabolic machinery)?
In order to address this question, we consider a self-replicating system composed of the gene expression machinery ($R$) and the metabolic machinery ($M$).
The system, shown schematically in Fig.~\ref{fig:self_replicator}, is thus defined by two macroreactions which are conveniently written as:
\begin{equation}\label{eq:reactions}
\begin{aligned}
S &\overset{V_M}{\longrightarrow} P, \\
P &\overset{V_R}{\longrightarrow} \alpha R + (1-\alpha) M.
\end{aligned}
\end{equation}
The first reaction, catalyzed by $M$, converts external substrates ($S$) into precursor metabolites ($P$).
The second reaction, catalyzed by $R$, converts precursors into macromolecules ($R$ and $M$).
The resource allocation parameter $\alpha \in [0,1]$ defines the proportion of precursor mass used for making gene expression machinery as compared to metabolic machinery.
We will interchangeably use the symbols $M$, $R$, $S$, and $P$ for the components of the replicators themselves and their total mass [g].
We will denote the rates at which the macroreactions occur by $V_R$ and $V_M$ [g~h\textsuperscript{-1}].

@@

\subsection{A near-optimal feedback control strategy that exploits information on the imbalance between precursors and the gene expression machinery}
\label{sec:strategies}

@@

As described in the \textit{Results} section, we are interested in control strategies satisfying the following conditions:
\begin{description}
\item[(C1)] The control laws are static functions of the system variables (as opposed to, for instance, functions that depend on derivatives or integrals of the variables).
\item[(C2)] For any given constant environment $E_M$, they drive the self-rep\-li\-cat\-or system towards a unique stable steady state that is not trivial, \textit{i.e.}, with nonzero growth rate.
\item[(C3)] This steady state corresponds to the optimal steady state $(\hat{p}_{opt}^*, \linebreak \hat{r}_{opt}^*)$, allowing growth at the maximal rate $\mu^*_{opt}$.
\end{description}
It can be directly verified from the functions $f$, $g$, and $h$ defining the nutrient-only, precursor-only, and on-off control strategies (Eqs~\ref{eq:f}, \ref{eq:precursorstrategy}, and \ref{eq:stratswitch}) that they are indeed static functions of the system variables (or the system input, in the case of the nutrient-only strategy). 
Here we show that the other two conditions are also satisfied for all three strategies.

@@

The full specification of the precursor-only strategy demands an expression for the function $g$ in Eq.~\ref{eq:precursorstrategy}.
Recall that Eq.~\ref{eq:alpha_mu_optimal_p} defines $\alpha_{opt}^*$ in terms of the precursor concentration $\hat{p}_{opt}^*$, which leads us to propose the following function $g$:
\begin{equation}
\label{eq:meth_g}
g(\hat{p}) = \frac{\hat{p}}{\hat{p} + \frac{K}{K + \hat{p}}(1+\hat{p})}.
\end{equation}
As shown in \ref{S1_Text} by computing the Jacobian, the system given by \linebreak Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} and \ref{eq:meth_g} has a single nontrivial stable steady state for any environment $E_M$ (Condition~C2).
Moreover, Eq.~\ref{eq:meth_g} guarantees this steady state to be optimal (Condition~C3).
This can be seen by noting that at steady state, $d\hat{r}/dt=0$ implies $\hat{r}^*=g(\hat{p}^*)$ (Eq.~\ref{eq:rdefnondim}).
In order for the self-replicator to attain a maximal growth rate at steady rate, Eq.~\ref{eq:alpha_mu_optimal_p} needs to be satified, which is the case for the above choice of the function $g$.
Like for $f$, Eq.~\ref{eq:meth_g} is the only choice for $g$ satisfying C1-C3.
\ref{S1_Fig} shows a plot of $g(\hat{p})$ together with a biologically plausible Hill approximation (Eq.~\ref{eq:HillApprox}).

The on-off control strategy is defined in Eq.~\ref{eq:stratswitch} and repeated below:
\begin{equation}
h(\hat{p}, \hat{r}) = 
\begin{cases}
0, \ \textrm{if} \ \hat{r} > g(\hat{p}),\\
1, \ \textrm{if} \ \hat{r} < g(\hat{p}), \\
\alpha_{opt}^*, \ \textrm{if} \ (\hat{p},\hat{r})=(\hat{p}_{opt}^*,\hat{r}_{opt}^*).
\end{cases}
\end{equation}
This strategy drives the system to a single steady state, because the $\hat{p}$-nullcline crosses the function $g(\hat{p})$ only once, as shown graphically in \linebreak Fig.~\ref{fig:onoffresults}\textit{A}.
In \ref{S1_Text} we argue that this steady state is stable, by taking into account so-called sliding modes on the switching curve~\cite{filippov_differential_1988} (Condition~C2).
Moreover, the steady state coincides with the optimal steady state $(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$ by construction, so that Condition~C3 is satisfied as well.
Fig~\ref{fig:ppGppsurface}\textit{A} shows a plot of $h(\hat{p}, \hat{r})$.

@@

In this section, we derive the main results for the functions $f$, $g$, and $h$ defining the nutrient-only, precursor-only, and on-off control strategies.
For each of these, we prove that the Conditions C1, C2 and C3 from the \textit{Methods} section are satisfied, which we repeat here for clarity:
\begin{description}
\item[(C1)] The control laws are static functions of the system variables (as opposed to, for instance, functions that depend on derivatives or integrals of the variables).
\item[(C2)] For any given constant environment $E_M$, they drive the self-rep\-li\-cat\-or system towards a unique stable steady state that is not trivial, \textit{i.e.}, with nonzero growth rate.
\item[(C3)] This steady state corresponds to the optimal steady state $(\hat{p}_{opt}^*, \linebreak \hat{r}_{opt}^*)$, allowing growth at the maximal rate $\mu^*_{opt}$.
\end{description}

@@

Ref.~\cite{uter_longrange_2004} provides a typical value for such a reaction, catalyzed by glu\-ta\-min\-yl-tRNA synthetase: $k_{\textit{cat,GlnRS}} = 3.2$~s\textsuperscript{-1}, indicating that on average 3.2 glutaminyl-tRNA molecules are produced per glutaminyl-tRNA synthetase molecule per second. 
After conversion to mass units using molar weight from~\cite{freist_glutaminyltrna_1997}, this yields
\[
k_{\textit{cat,GlnRS}} = \frac{3.2 \cdot 147}{64.4 \cdot 10^ 3} \approx 10^{-3} \, \text{g of glutaminyl-tRNA} \cdot \text{g of enzyme}^{-1} \cdot \text{s}^{-1}.
\]
We therefore take
\[
k_M \approx 3.6 \text{ h}^{-1},
\]
and thus obtain an upper bound for $e_M$ in our simulations.

@@

An extremal trajectory is a quadruplet $(\hat{x}(\cdot),\lambda(\cdot),\lambda_0,\alpha(\cdot))$ satisfying \linebreak Eqs~\ref{eq:model}-\ref{eq:PMP}. The extremal is said to be normal (resp. abnormal) if $\lambda_0<0$ (resp. $\lambda_0=0$). In the normal case, we normalize the adjoint vector so that $\lambda_0=-1$.

@@

\subsection{S1 Table -- Parameter values of the self-rep\-li\-cat\-or model}
\manuallabel{S1_Table}{S1~Table}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{./Fig/plot_strategies}
\caption{
\textbf{Simple control strategies for the self-rep\-li\-cat\-or of bacterial growth.}
\textit{A:} Nutrient-only strategy: $\alpha = f(E_M)$.
The dashed, black curve is the (unique) strategy driving the system exactly to the optimal steady state, that is, the state in which growth occurs at the maximal rate supported by $E_M$.
The function $f$ is defined by Eq.~\ref{eq:meth_f} in the \textit{Methods} section of the main text.
The solid, red curve is an approximation of this function by the simple Michaelis-Menten curve of Eq.~\ref{eq:MMApprox}, with $K_{mE} = 1.0$.
\textit{B:} Precursor-only strategy: $\alpha = g(\hat{p})$.
The dashed, black curve is the (unique) strategy driving the system exactly to the optimal steady state.
The function $g$ is defined by Eq.~\ref{eq:meth_g} in the \textit{Methods} section of the main text.
The solid, red curve is an approximation of this function by the simple Hill curve of Eq.~\ref{eq:HillApprox}, with $K_{mp} = 0.06$ and a cooperativity coefficient 2.}
\end{figure}

### 2017-04-05 abstract.tex
Some small frenchproofreading and correct overfull/underfull hbox when possible

Même si de vieilles données le suggèrent, un tel comportement \textit{bang-bang} n'a jamais été totalement confirmé expérimentalement pour la synthèse des ribosomes.
Nous mesurons donc l'abondance des ribosomes chez \textit{Escherichia coli} au niveau d'une cellule unique lors d'un changement brutal de milieu de culture.
En particulier, nous créons une souche de \textit{E.coli} sur laquelle un rapporteur fluorescent est attaché à l'une des sous-unités ribosomales, permettant ainsi leur quantification \textit{in vivo}.
Un appareillage micro-fluidique nous permet ensuite de contrôler en temps réel le milieu de croissance tout en observant individuellement chaque cellule fluorescente.
Nous développons une méthode basée sur le lissage de Kalman qui est capable de reconstruire la façon dont les ressources sont aiguillées vers la synthèse des ribosomes.
Même si ces résultats sont préliminaires, ils suggèrent que la concentration des ribosomes oscille après le changement d'environnement, ce qui rappelle une stratégie de type \textit{bang-bang}.

@@

\begin{resume-vulgaire}
Bactéries et autres micro-organismes se multiplient en utilisant la matière et l'énergie présentes dans leur environnement.
Les stratégies qui leur permettent d'optimiser ce processus ont été longuement étudiées, mais uniquement dans des environnements stables.
Ici, nous construisons un modèle simple de la croissance microbienne et utilisons la théorie du contrôle optimal pour déterminer la façon dont les cellules devraient agir lorsque leur environnement est soudainement modifié.
Nous mettons en évidence que la meilleure stratégie consiste à aiguiller toutes les ressources disponibles vers la production d'un seul composant de manière alternative, une prédiction que nous tentons également de confirmer de manière expérimentale.
En d'autres termes, dans un contexte dynamique, mieux vaut agir tel un interrupteur à bascule plutôt qu'un variateur de lumière.
Ces résultats pourraient suggérer de nouvelles façons d'optimiser les processus biotechnologiques de production de molécules d'intérêt.
\end{resume-vulgaire}

>>>>

Même si de vieilles données le suggèrent, un tel comportement \textit{bang-bang} n'a jamais été totalement confirmé expérimentalement pour la synthèse des ribosomes.
Nous mesurons donc l'abondance des ribosomes chez \textit{Escherichia coli} au niveau de cellules individuelles lors d'un changement brutal de milieu de culture.
En particulier, nous créons une souche de \textit{E.coli} sur laquelle un rapporteur fluorescent est attaché à l'une des sous-unités ribosomales, permettant ainsi leur quantification \textit{in vivo}.
Un appareillage micro-fluidique nous permet ensuite de contrôler en temps réel le milieu de croissance tout en observant individuellement chaque cellule fluorescente.
Nous développons une méthode basée sur le lissage de Kalman qui est capable de reconstruire la façon dont les ressources sont aiguillées vers la synthèse des ribosomes.
Même si ces résultats sont préliminaires, ils suggèrent que la concentration des ribosomes oscille après le changement d'environnement, ce qui rappelle une stratégie de type \textit{bang-bang}.

@@

\begin{resume-vulgaire}
Bactéries et autres micro-organismes se multiplient en utilisant la matière et l'énergie présentes dans leur environnement.
Les stratégies qui leur permettent d'optimiser ce processus ont été longuement étudiées, mais uniquement dans des environnements stables.
Ici, nous construisons un modèle simple de la croissance microbienne et utilisons la théorie du contrôle optimal pour déterminer la façon dont les cellules devraient agir lorsque leur environnement est soudainement modifié.
Nous mettons en évidence que la meilleure stratégie consiste à aiguiller toutes les ressources disponibles vers la production d'un seul composant de manière alternative, une prédiction que nous tentons également de confirmer de manière expérimentale.
Ces résultats pourraient suggérer de nouvelles façons d'optimiser les processus biotechnologiques de production de molécules d'intérêt.
\end{resume-vulgaire}
